Cartogram Mapping and its Application to Cancer Data Visualization

### 2019-09-24 review_article.Rmd
spell check for US and missing years added to references

---
title: "Cartogram Mapping and its Application to Cancer Data Visualization"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
bibliography: reviewbib.bib
output:
  word_document:
    toc: yes
    toc_depth: '3'
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
link-citations: yes
fontsize: 11pt
biblio-style: vancouver
yolo: yes
---

@@

- Future steps should consider applications not currently utilised for cancer outcome displays, where their use may improve decision making
- needs to have a clear relevance to spatial patterns in cancer epidemiology
- presents alternative ways of visualizing those spatial patterns, or the uncertainty in those patterns
- a more applied-focused paper (e.g. lots of cool images), rather than theoretical (e.g. lots of formula).
-->

@@

# Introduction
<!-- What data are we working with, why, what are the purposes -->
Cancer statistics are usually delivered as an aggregated value for a geopolitical area. Presenting these statistics requires transforming individual observations into aggregations of communities as geographical units, in large part for privacy protection or political and policy purposes.  The information could be as simple as counts per area (e.g. state, province, local government area, post/zip code). Counts alone is not sufficient to compare different areas, because the populations of different areas are likely different. In this case, the counts data needs to be merged with population data to appropriately calibrate it to incidence, for example, rate per 100000 people.  This type of data is collected on a routine basis for public health purposes, and may be made available to the general public as a service to the community. The task, then, is to examine what are the usual ways to communicate cancer statistics, to the public, are there alternative approaches, and what are the pros and cons of these choices.

<!-- Ways to visualize, whats our hypothesis -->
A common approach to communicate cancer statistics, is to display statistics on a map. Using a choropleth map: the statistic is mapped to color and the geographic region is filled with this color. The viewer would then be able to examine the spatial distribution of the disease incidence, where there is a trend in longitude or latitude, or rural vs urban, or coastal vs inland, or even specific hot spots of the disease. Visualizing diseases on maps is often the first step in exploratory spatial data analysis and effectively helps in the formulation of hypotheses [@MTMSIH]. Disease maps help to present geographic patterns that may be overlooked in a table, obscuring the geospatially related statistics [@SAMGIS].
By providing a visual representation of cancer outcomes, geographic patterns of disease are able to be identified and effectively addressed with public health policy and actions.
@SE recognizes one of the key challenges with mapping spatial patterns of disease is the design of visualizations.
This paper addresses the visualization techniques and their applications to cancer statistics. Highlighting the differences and historic use of these displays.

<!-- Structure of the paper -->
The paper is organised as follows. The next section describes the choropleth map which is the common approach to disease maps. Section \ref{ch:public} surveys atlases in use today. Section \ref{ch:alternatives} describes an alternative display, the cartogram what may be useful for countries that have heterogeneously sized geographic units. The pros and cons of these approaches is discussed in Section \ref{ch:compare}. Disease maps are more useful when made interactive, and common options are described in Section \ref{ch:interacting}, along with a discussion of benefits and disadvantages. The last section summarizes the paper and discusses future directions.

@@

<!-- how do we use choropleth's -->
Identifying and explaining spatial structures, patterns, and processes involves considering the individuals in communities and organizing communities into representable units [@SAMGIS]. In Figure \ref{fig:choroCRS}, a west to east spatial trend of increasing rates, can be seen. There is also a spatial outlier -- Utah has a noticeably lower rate that its neighbours. Also Kentucky has a noticeably high rate, and Maine also has a higher rate than its neighbours. There is something of a cluster of higher rates around the tobacco states.

@@

```{r choropleth-grid-create, message=FALSE, warning = FALSE, fig.cap = "Publicly available choropleth cancer maps, published between 2010 and 2015.", out.width='100%', fig.align='center', fig.width=12, fig.height=10}
library(cowplot)

@@

The atlas of Cancer in Queensland [@DACQ] focused on highlighting the difference in experience for those living in rural and disadvantaged areas. They used Standardized Incidence Ratios modeled via Bayesian methods, and explored the information presented in a range of selected atlases. These atlases provided Incidence, Survival and Mortality rates to the public on Areal or Isopleth map bases.
The presentation of these rates considers not only the shapes of the areas, but also appropriate choices of color blind friendly color schemes, and categories of the values to highlight significantly different areas.

Cancer maps are effective tools for communicating to wide range of audiences, including the general public and others not trained in statistical analyses.
These visualizations enable non-expert audiences to interpret the outputs of sophisticated statistical analyses. Cruickshank (1947) as cited by @DMAHP, discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly, cancer maps utilised to communicate to the public and other non-expert audiences are choropleths.

@@

@roberts2019communication identified 33 publicly available cancer atlases, published between January 2010 and November 2015. All of these use choropleth maps. All but one of these were published by non-commercial organisations, including not-for-profits, government, research organisations, advocacy groups or government funded partnerships. The use of choropleths within the public domain mirrors the heavy use of choropleth maps within the research literature, discussed above. 
The cancer atlases identified by @roberts2019communication covered geographies from all around the world, most focused on single nations. Figure \ref{fig:choropleth-grid-create} displays a global (b), national (a,d,e,f) and state (c,g) choropleths. The sections below provide details on each of the maps within Figure \ref{fig:choropleth-grid-create}.

@@

Figure 2.a contains an image from the *[The Environmental and Health Atlas of England and Wales](http://www.envhealthatlas.co.uk/eha/Breast/)*. This map shows the relative risk for women developing lung cancer in England and Wales in 2010. The cancer data used to generate this map came from Office for National Statistics (ONS) (England) and from the Welsh Cancer Intelligence and Surveillance Unit (WCISU).

@@

The map seen in Figure 2.b is from the *[Globocan 2012: Estimated  Cancer Incidence, Mortality and Prevalence](http://globocan.iarc.fr/Pages/Map.aspx)*. This global map shows age standardized incidence rates (per 100,000) for all invasive cancers for both men and women, aggregated at a national level for 2018. This map is published by the [World Health Organisation's International Agency for Research on Cancer](https://www.iarc.fr/). Data was sourced from cancer registries in each country, contributing registries can be seen in the supplementary material on the cancer atlas website.

@@

Figure 2.c shows an extract from the *[The Atlas of Cancer in QLD](https://cancerqld.org.au/research/queensland-cancer-statistics/queensland-cancer-atlas/)*. This map was published by the Queensland Cancer Council and shows the relative incidence ratio of lung cancer in males in the state of QLD within Australia based on data from 1998 to 2007. Data to generate this map was sourced from the *[Queensland Cancer Registry](https://cancerqld.org.au/research/cancer-registries/queensland-cancer-registry/)*.

@@

One of the concerns of adding too much information to a map is the fear of cognitive overload [@mcgranaghan1993cartographic] in which the the user reaches an information threshold, beyond which will not be able to make sense of the information. These concerns are not unfounded, complexity and density of representation methods appear to overwhelm novice decision makers, while experts are able to use the detail more readily when making-decisions [@cliburn2002design]. Interactivity is a design feature within modern mapping methods that can be used to incorporate additional information and complexity without overloading the user.

@@

The interactivity of modern mapping methods enable supplementary information to be incorporated without cluttering the screen.
Additionally, interactivity allows the user to toggle between different variables, map views or multiple realizations of possible future scenarios [@goodchild1994introduction]. Thus providing additional mechanisms for the users comprehension as well as the uncertainty of the available information (@maceachren1992visualizing; @van1994visualization).

@@

While atlases are often used to describe differences between areas, statistics may be displayed at different levels of aggregation. Global health statistics can be aggregated to administrative and arbitrarily defined regions, such as those used by the World Health Organisation and the United Nations [@IARC_3].
World atlases can allow for displays of data aggregated into continents, countries, states, provinces and congressional districts [@USCS].

@@

It is extremely likely that each population area will have a different number of people. The distribution of the population residing in all areas may also be communicated in a table or histogram display [@NICR_1].
Atlases have the opportunity to connect the population to the land available to them by communicating population density.
Atlases can also connect the population to the land available to them by communicating population density.

### Statistical uncertainty
Additional statistics that accompany an atlas often include a measure of the statistical uncertainty surrounding the statistics presented in a choropleth.
In the review of atlases in the public domain, atlases were considered to report uncertainty to the non-expert user if they included a measure of statistical uncertainty either within or alongside the map [@roberts2019communication]. 
The maps considered used standard and well known measures including credible intervals and standard deviation, statistical significance, box plots and distributions. Other methods involve providing adjacent maps or overlapping maps with symbols [@VSSDCUC].
The maps employing uncertainty ranged from static documents or info-graphics, to interactive online visualizations. Communicating the statistical uncertainty associated with the estimates occurs using confidence intervals (CI), credible intervals (CrI), statistical significance levels, box plots, distribution plots, and reporting sample size and standard deviations.
Close to half of the atlases identified (42%, n=14) included some measure of uncertainty. The most common measure used to represent uncertainty were credible or confidence intervals (CIs).

@@

### Socio-economic indicators
Socio-economic indicators can explain how the experience of cancer prevalence varies for various members of a society.
These indicators include unemployment rates, poverty rates, remoteness, and education levels achieved though, only a few atlases also explored the impact of rurality on cancer rates. 
These rates may also be explored as percentages above or below the mean or median value for the set of spatial areas.
The Human Development Index can be used to understand the socio-economic experience of a community, as well as Income levels which can be sourced from the World Bank list of economies [@IARC_3]. 
The areas are often ranked and allocated to quintiles, each quintile can be presented as categories describing the ranking.

One of the concerns of adding too much information to a map is the fear of cognitive overload [@mcgranaghan1993cartographic] in which the the user reaches an information threshold, beyond which will not be able to make sense of the information. These concerns are not unfounded, complexity and density of representation methods appear to overwhelm novice decision makers, while experts are able to use the detail more readily when making-decisions [@cliburn2002design]. Interactivity is a design feature within modern mapping methods that can be used to incorporate additional information and complexity without overloading the user.

@@

Choropleths are uni-variate displays, they do not inherently support multi-dimensionality.
This makes it difficult to pair demographic, environmental and other factors with spatial distributions.
@MMST present linked micromap plots as a solution to the issue of multi-dimensionality to visually link geographic and statistical data.
This design is widely used in online atlas displays, it groups areas based on their value for one variable, and uses  additional columns to provide displays that contrast the areas in each group by other variables.
The display juxtaposes of choropleth maps and statistical plots;
it shows one map per group of the key separating variable, in a row with each additional statistical plot.
Linked micromaps predominantly utilise the choropleth map for displays of spatial distributions.
It is one of several alternative displays that allow maps to become bi-variate displays.

@@

<!-- What is a cartogram? -->
Choropleths imply uniformity of data across the geographic space, when considering population densities it is extremely unlikely to be uniform [@BCM].
A cartogram displays abstracted geographic space, with the intention of improving the presentation of the statistic of interest. For a single variable of interest, each map area is changed to emphasise the distribution by representing the corresponding value, in comparison to the value of the other areas [@ACCAC].
Changes in the map base are implemented by altering the boundaries, and therefore shapes, of individual areas.

@@

<!-- There are many alternative views of the world-->
<!-- Tobler map projections-->
Choropleths may be considered true topological maps, however, if the land mass displayed covers enough of the globe, there must be a transformation or distortion to display the land in 2D [@HTLWM]. The amount of distortion is related to the distance covered by the landmass displayed @GAMP.
World map projections reflect the frequent perspectives used to view the earth. Choropleth maps will always be distorted if they cover enough of the globe, just like photographs of the globe from space. Choropleth creation requires choosing a map projection that shows a favourable distortion of the geography for presenting the set of spatial information. Selecting a display can prevent misinterpretation of global statistics, as global maps face the challenge of equitable displays of land mass on maps [@RSCW].
If the statistic presented on the map base relies on physical distance and is influenced by the topology there is no transformation needed, beyond choosing a reasonable projection.

@@

<!-- Why transform?-->
<!-- Common variables used to create cartograms, e.g. population, mortality -->
Cartograms provide an alternative visualization method for statistical and geographical information.
The key difference between a choropleth and a cartogram is the desirable augmentation of the size, shape or distance of geographical areas [@ACTUC]. 
@HTLWM suggests that white lies may be employed to create useful displays and map creators have the ability to draw lines that may distort the geometry and suppress features and it is easy for the average person to disregard the impact of transformations used to create cartograms.
Cartograms may be seen as an extension of map transformations and projections. The favourable distortion is proportional to a value other other than actual earth size area [@NAC].
A disadvantage of the conventional map is that sparsely populated rural areas may be emphasized, whereas the areas representing cities are very small, making interpretation of spatial patterns very difficult.
The distortion of a cartogram accounts for the population density, preventing it from obscuring the spatial patterns [@TAAM]. 
The spatial transformation of map regions relative to the data emphasises the data distribution instead of land size [@CBATCC].
When visualizing population statistics, @ACTUC considers this equitable representation design 'more socially just', or honest [@NISCC], giving due attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. 
@HEDP suggests that 'cancer occurs in people, not in geographical areas' and @CTTMB believe that spatial socio-economic data, like cancer rates, are best presented on a cartogram for urban areas as the population map base avoids allocating 'undue prominence' to rural areas. 
@MTMSIH encourage the use of cartograms to highlight small areas and uncover local-level inequalities.

@@

The creation of cartograms was historically in the hands of professional cartographers [@CD]. 
Early approaches including John Hunter and Jonathan Young (1968) and Durham's wooden tile method, Skoda and Robertson's (1972) steel ball bearing approach and Tobler's (1973) computer programs [@ACTUC]. 
Geographical information systems allowed map users, and researchers to implement their own cartograms, but these systems are utilised depending on ‘the effectiveness, efficiency, and satisfaction of the map products (Nielsen 1994), [@CD].
@HEDP discusses the impact of electronic computer-assisted techniques.

@@

|Map display |Details| 
|:--------------------|:------------------------------------------------|  
|a. contiguous cartogram| Each stated shape has been distorted according to the population of the state in 2015. The state of California has become much larger due to it's large population density. This draws attention to the densely populated North East region, and detracts from the less populated Mid West.|
|b. Non - Contiguous| The geographic shape of the states has been maintained, but the size has altered according to the population of the state in 2015. The state of California has remained closer to it's original size than it's surrounding states. The North East states have remained closer to their geographical size, in the case of Massachusetts and Connecticut. This draws attention to the densely populated North East region, and the sparse Mid West.|
|Dorling| The states have been represented by a circle, but the size was determined by the population of the state in 2015. The North East states remain closer to their neighbours, and may be displaced from their geographic location. The sparsity of the population in the Mid West is highlight by the distance between the circles, located at the geographic centroids.|
|Hexagon Tessellation| Each state is represented by a hexagon of equal size. The neighbouring states are easily contrasted, however the north east regions has been displaced from their geographic location. The sparsity of the population in the Mid West is highlight by the light yellow color, the age adjusted rate in Kentucky is the darkest and its neighbours are similar.|

@@

<!-- What is a contiguous cartogram -->
<!-- Intentionally preserve neighbours -->

@@

The groundwork for this approach had been laid in the mid 1930's by Raisz, and rectangular cartograms provide dramatic comparisons and are especially useful for correcting misconceptions communicated by geographic maps. 
@TFYCC quotes the official definition of Value-Area Cartograms, the simplistic displays which represent each area as a single rectangle, sized according to the value of the statistic.
This rectangular display also allows for tiling, where geographic neighbours placed in suitable relative positions also share borders, however contiguity may be sacrificed [@CDWCS].
Rectangular cartograms allow for bi-variate displays, population can be effectively communicated by the size of each rectangular, and a second variable can be communicated using color [@ORC].

@@

A simple tile map presents the areas in a tessellated grid display, where geographic neighbours are found next to each other, with some necessary displacement employed for regions with more than four neighbours.
These tiles may be labelled or colored to represent a value.
Tile maps may be difficult to create, they are best created manually, with additional time and care required as the amount of geographic areas to include increases.
@MDAC define the term 'mosaic cartograms' for hexagonal tile displays, where the number of tiles for each area can be used to communicate the statistic of regions. 
The complexity of the boundaries can be adjusted in the resulting display, as the size of the tiles used allows a trade-off be made between boundary complexity and simplicity.

@@

<!-- Choropleths, true topology, geographic distance, good for showing geographic importance -->
To choose an appropriate map display, the map creator must consider the intended user, and message the map will communicate.
Choropleth displays utilise more traditional cartographic methods, they are usually true to the topography, displaying familiar boundaries of countries, states or administrative areas. Highlighting the geographic distribution, and differences between equal spatial units.
When there is a relationship between the variable being mapped and geographical features, a choropleth map would pair this information visually.
Choropleth methods require a decision to be made about the projection of the display, in this case aim to find a map projection that gives the least distorted representation of the geography [@UMPT].
However a choropleth is limited to a uni-variate display, presenting a spatial distribution by filling each area with a color.
These displays favor geographically large regions over small regions, though inset displays are often used to zoom in on a geographic area containing many small regions, this is seen in Figure \ref{fig:choropleth-grid-create} c of the state of Queensland, where the capital city Brisbane is highlighted by the inset map. 
Users can identify the areas relevant to them by the familiar boundaries, and can easily contrast the experience of geographic neighbours. The map is reusable and fit for many purposes.

<!-- Cartograms, sacrifice geography, focus on the outliers/ relationships in data space -->
Cartographers were historically the creators of geographic displays, and epidemiologists utilised the maps that had been prepared in advance.
Cartograms with the population variable used to distortion the geographic map regions and boundaries are density equalising maps, that reduce the visual prominence of large, but low-population areas [@VSSDCUC]. 
Cartogram displays incorporate the statistics and population into the design of the display,
shifting and sometimes sacrificing familiar boundaries to draw attention to large outliers in the data space or population density.
The unusual shapes are not longer familiar and may be difficult to compare. This can prevent effective comparison of the statistic. Cartograms also favour extreme statistics in the positive direction. Small extreme values are often lost due to the resulting small sized areas the do not draw user's attention.
The change in the map based on one variable provides the opportunity for bi-variate displays, using the change in area for one variable, and color for another.
The difference between the familiar map display helps highlight the impact of the disease on communities.

>>>>

---
title: "Cartogram Mapping and its Application to Cancer Data Visualization"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
bibliography: reviewbib.bib
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  word_document:
    toc: yes
    toc_depth: '3'
link-citations: yes
fontsize: 11pt
biblio-style: vancouver
yolo: yes
---

@@

- Future steps should consider applications not currently utilized for cancer outcome displays, where their use may improve decision making
- needs to have a clear relevance to spatial patterns in cancer epidemiology
- presents alternative ways of visualizing those spatial patterns, or the uncertainty in those patterns
- a more applied-focused paper (e.g. lots of cool images), rather than theoretical (e.g. lots of formula).
-->

@@

# Introduction
<!-- What data are we working with, why, what are the purposes -->
Cancer statistics are usually delivered as an aggregated value for a geopolitical area. Presenting these statistics requires transforming individual observations into aggregations of communities as geographical units, in large part for privacy protection or political and policy purposes.  The information could be as simple as counts per area (e.g. state, province, local government area, post/zip code). Counts alone are not sufficient to compare areas, because the populations of areas are all different. In this case, the counts data needs to be merged with population data to appropriately calibrate it to incidence, for example, rate per 100000 people.  This type of data is collected on a routine basis for public health purposes, and may be made available to the general public as a service to the community. The task, then, is to examine what are the usual ways to communicate cancer statistics, to the public, are there alternative approaches, and what are the pros and cons of these choices.

<!-- Ways to visualize, whats our hypothesis -->
A common approach to communicate cancer statistics, is to display statistics on a map. Using a choropleth map: the statistic is mapped to color and the geographic region is filled with this color. The viewer would then be able to examine the spatial distribution of the disease incidence, where there is a trend in longitude or latitude, or rural vs urban, or coastal vs inland, or even specific hot spots of the disease. Visualizing diseases on maps is often the first step in exploratory spatial data analysis and effectively helps in the formulation of hypotheses [@MTMSIH]. Disease maps help to present geographic patterns that may be overlooked in a table, obscuring the geospatially related statistics [@SAMGIS].
By providing a visual representation of cancer outcomes, geographic patterns of disease may be identified and effectively addressed with public health policy and actions.
@SE recognizes one of the key challenges with mapping spatial patterns of disease is the design of visualizations.
This paper addresses the visualization techniques and their applications to cancer statistics. Highlighting the differences and historic use of these displays.

<!-- Structure of the paper -->
The paper is organized as follows. The next section describes the choropleth map which is the common approach to disease maps. Section \ref{ch:public} surveys atlases in use today. Section \ref{ch:alternatives} describes an alternative display, the cartogram what may be useful for countries that have heterogeneously sized geographic units. The pros and cons of these approaches is discussed in Section \ref{ch:compare}. Disease maps are more useful when made interactive, and common options are described in Section \ref{ch:interacting}, along with a discussion of benefits and disadvantages. The last section summarizes the paper and discusses future directions.

@@

<!-- how do we use choropleth's -->
Identifying and explaining spatial structures, patterns, and processes involves considering the individuals in communities and organizing communities into representable units [@SAMGIS]. In Figure \ref{fig:choroCRS}, a west to east spatial trend of increasing rates, can be seen. There is also a spatial outlier -- Utah has a noticeably lower rate that its neighbors. Also Kentucky has a noticeably high rate, and Maine also has a higher rate than its neighbors. There is something of a cluster of higher rates around the tobacco states.

@@

```{r choropleth-grid-create, message=FALSE, warning = FALSE, fig.cap = "Publicly available choropleth cancer maps published between 2010 and 2015.", out.width='100%', fig.align='center', fig.width=12, fig.height=10}
library(cowplot)

@@

Cancer maps are effective tools for communicating to wide range of audiences, including the general public and others not trained in statistical analyses.
These visualizations enable non-expert audiences to interpret the outputs of sophisticated statistical analyses. Cruickshank (1947) as cited by @DMAHP, discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly, cancer maps utilized to communicate to the public and other non-expert audiences are choropleths.  These atlases provided Incidence, Survival and Mortality rates to the public on areal or isopleth map bases.

The Atlas of Cancer in Queensland [@DACQ] focused on highlighting the difference in experience for those living in rural and disadvantaged areas, the Standardized Incidence Ratios (SIR) were modeled via Bayesian methods.
The presentation of these rates considers not only the shapes of the areas, but also appropriate choices of color blind friendly color schemes, and categories of the values to highlight significantly different areas.

@@

@roberts2019communication identified 33 publicly available cancer atlases, published between January 2010 and November 2015. All of these use choropleth maps. All but one of these were published by non-commercial organizations, including not-for-profits, government, research organizations, advocacy groups or government funded partnerships. The use of choropleths within the public domain mirrors the heavy use of choropleth maps within the research literature, discussed above. 
The cancer atlases identified by @roberts2019communication covered geographies from all around the world, most focused on single nations. Figure \ref{fig:choropleth-grid-create} displays a global (b), national (a, d, e, f) and state (c, g) choropleths. The sections below provide details on each of the maps within Figure \ref{fig:choropleth-grid-create}.

@@

Figure 2.a contains an image from *[The Environmental and Health Atlas of England and Wales](http://www.envhealthatlas.co.uk/eha/Breast/)*. This map shows the relative risk for women developing lung cancer in England and Wales in 2010. The cancer data used to generate this map came from Office for National Statistics (ONS) (England) and from the Welsh Cancer Intelligence and Surveillance Unit (WCISU).

@@

The map seen in Figure 2.b is from the *[Globocan 2012: Estimated  Cancer Incidence, Mortality and Prevalence](http://globocan.iarc.fr/Pages/Map.aspx)*. This global map shows age standardized incidence rates (per 100,000) for all invasive cancers for both men and women, aggregated at a national level for 2018. This map is published by the [World Health organization's International Agency for Research on Cancer](https://www.iarc.fr/). Data was sourced from cancer registries in each country, contributing registries can be seen in the supplementary material on the cancer atlas website.

@@

Figure 2.c shows an extract from  *[The Atlas of Cancer in QLD](https://cancerqld.org.au/research/queensland-cancer-statistics/queensland-cancer-atlas/)*. This map was published by the Queensland Cancer Council and shows the relative incidence ratio of lung cancer in males in the state of QLD within Australia based on data from 1998 to 2007. Data to generate this map was sourced from the *[Queensland Cancer Registry](https://cancerqld.org.au/research/cancer-registries/queensland-cancer-registry/)*.

@@

One of the concerns of adding too much information to a map is the fear of cognitive overload [@mcgranaghan1993cartographic] in which the user reaches an information threshold, beyond which will not be able to make sense of the information. These concerns are not unfounded, complexity and density of representation methods appear to overwhelm novice decision makers, while experts are able to use the detail more readily when making-decisions [@cliburn2002design]. Interactivity is a design feature within modern mapping methods that can be used to incorporate additional information and complexity without overloading the user.

@@

The interactivity of modern mapping methods enables supplementary information to be incorporated without cluttering the screen.
Additionally, interactivity allows the user to toggle between different variables, map views or multiple realizations of possible future scenarios [@goodchild1994introduction]. Thus providing additional mechanisms for the users comprehension as well as the uncertainty of the available information (@maceachren1992visualizing; @van1994visualization).

@@

While atlases are often used to describe differences between areas, statistics may be displayed at different levels of aggregation. Global health statistics can be aggregated to administrative and arbitrarily defined regions, such as those used by the World Health organization and the United Nations [@IARC_3].
World atlases can allow for displays of data aggregated into continents, countries, states, provinces and congressional districts [@USCS].

@@

It is extremely likely that each population area will have a different number of people. The distribution of the population residing in all areas may also be communicated in a table or histogram display [@NICR_1].
Atlases can connect the population to the land available to them by communicating population density.
Atlases can also connect the population to the land available to them by communicating population density.

### Statistical uncertainty
Additional statistics that accompany an atlas often include a measure of the statistical uncertainty surrounding the statistics presented in a choropleth.
In the review of atlases in the public domain, atlases were considered to report uncertainty to the non-expert user if they included a measure of statistical uncertainty either within or alongside the map [@roberts2019communication]. 
The maps considered used standard and well-known measures including credible intervals and standard deviation, statistical significance, box plots and distributions. Other methods involve providing adjacent maps or overlapping maps with symbols [@VSSDCUC].
The maps employing uncertainty ranged from static documents or infographics, to interactive online visualizations. Communicating the statistical uncertainty associated with the estimates occurs using confidence intervals (CI), credible intervals (CrI), statistical significance levels, box plots, distribution plots, and reporting sample size and standard deviations.
Close to half of the atlases identified (42%, n=14) included some measure of uncertainty. The most common measure used to represent uncertainty were credible or confidence intervals (CIs).

@@

### Socio-economic indicators
Socio-economic indicators can explain how the experience of cancer prevalence varies for various members of a society.
These indicators include unemployment rates, poverty rates, remoteness, and education levels achieved though, only a few atlases also explored the impact of rurality on cancer rates. 
These rates may also be explored as percentages above or below the mean or median value for the set of spatial areas.
The Human Development Index can be used to understand the socio-economic experience of a community, as well as Income levels which can be sourced from the World Bank list of economies [@IARC_3]. 
The areas are often ranked and allocated to quintiles; each quintile can be presented as categories describing the ranking.

One of the concerns of adding too much information to a map is the fear of cognitive overload [@mcgranaghan1993cartographic] in which the user reaches an information threshold, beyond which will not be able to make sense of the information. These concerns are not unfounded, complexity and density of representation methods appear to overwhelm novice decision makers, while experts are able to use the detail more readily when making-decisions [@cliburn2002design]. Interactivity is a design feature within modern mapping methods that can be used to incorporate additional information and complexity without overloading the user.

@@

Choropleths are uni-variate displays, they do not inherently support multi-dimensionality.
This makes it difficult to pair demographic, environmental and other factors with spatial distributions.
@MMST present linked micromap plots as a solution to the issue of multi-dimensionality to visually link geographic and statistical data.
This design is widely used in online atlas displays, it groups areas based on their value for one variable, and uses  additional columns to provide displays that contrast the areas in each group by other variables.
The display juxtaposes of choropleth maps and statistical plots;
it shows one map per group of the key separating variable, in a row with each additional statistical plot.
Linked micromaps predominantly utilize the choropleth map for displays of spatial distributions.
It is one of several alternative displays that allow maps to become bi-variate displays.

@@

<!-- What is a cartogram? -->
Choropleths imply uniformity of data across the geographic space however population densities are extremely unlikely to be uniform [@BCM].
A cartogram displays abstracted geographic space, with the intention of improving the presentation of the statistic of interest. For a single variable of interest, each map area is changed to emphasize the distribution by representing the corresponding value, in comparison to the value of the other areas [@ACCAC].
Changes in the map base are implemented by altering the boundaries, and therefore shapes, of individual areas.

@@

<!-- There are many alternative views of the world-->
<!-- Tobler map projections-->
Choropleths may be considered true topological maps, however, if the land mass displayed covers enough of the globe, there must be a transformation or distortion to display the land in 2D [@HTLWM]. The amount of distortion is related to the distance covered by the landmass displayed @GAMP.
World map projections reflect the frequent perspectives used to view the earth. Choropleth maps will always be distorted if they cover enough of the globe, just like photographs of the globe from space. Choropleth creation requires choosing a map projection that shows a favorable distortion of the geography for presenting the set of spatial information. Selecting a display can prevent misinterpretation of global statistics, as global maps face the challenge of equitable displays of land mass on maps [@RSCW].
If the statistic presented on the map base relies on physical distance and is influenced by the topology there is no transformation needed, beyond choosing a reasonable projection.

@@

<!-- Why transform?-->
<!-- Common variables used to create cartograms, e.g. population, mortality -->
Cartograms provide an alternative visualization method for statistical and geographical information.
The key difference between a choropleth and a cartogram is the desirable augmentation of the size, shape or distance of geographical areas [@ACTUC]. 
@HTLWM suggests that white lies may be employed to create useful displays and map creators have the ability to draw lines that may distort the geometry and suppress features and it is easy for the average person to disregard the impact of transformations used to create cartograms.
Cartograms may be seen as an extension of map transformations and projections. The favorable distortion is proportional to a value other other than actual earth size area [@NAC].
A disadvantage of the conventional map is that sparsely populated rural areas may be emphasized, whereas the areas representing cities are very small, making interpretation of spatial patterns very difficult.
The distortion of a cartogram accounts for the population density, preventing it from obscuring the spatial patterns [@TAAM]. 
The spatial transformation of map regions relative to the data emphasizes the data distribution instead of land size [@CBATCC].
When visualizing population statistics, @ACTUC considers this equitable representation design 'more socially just', or honest [@NISCC], giving due attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. 
@HEDP suggests that 'cancer occurs in people, not in geographical areas' and @CTTMB believe that spatial socio-economic data, like cancer rates, are best presented on a cartogram for urban areas as the population map base avoids allocating 'undue prominence' to rural areas. 
@MTMSIH encourage the use of cartograms to highlight small areas and uncover local-level inequalities.

@@

The creation of cartograms was historically in the hands of professional cartographers [@CD]. 
Early approaches including John Hunter and Jonathan Young (1968) and Durham's wooden tile method, Skoda and Robertson's (1972) steel ball bearing approach and Tobler's (1973) computer programs [@ACTUC]. 
Geographical information systems allowed map users, and researchers to implement their own cartograms, but these systems are utilized depending on ‘the effectiveness, efficiency, and satisfaction of the map products (Nielsen 1994), [@CD].
@HEDP discusses the impact of electronic computer-assisted techniques.

@@

|Map display |Details| 
|:--------------------|:------------------------------------------------|  
|a. contiguous cartogram| Each stated shape has been distorted according to the population of the state in 2015. The state of California has become much larger due to it's large population density. This draws attention to the densely populated North East region, and detracts from the less populated Mid West.|
|b. Non - Contiguous| The geographic shape of the states has been maintained, but the size has altered according to the population of the state in 2015. The state of California has remained closer to it's original size than it's surrounding states. The North East states have remained closer to their geographical size, in the case of Massachusetts and Connecticut. This draws attention to the densely populated North East region, and the sparse Mid West.|
|Dorling| The states have been represented by a circle, but the size was determined by the population of the state in 2015. The North East states remain closer to their neighbors, and may be displaced from their geographic location. The sparsity of the population in the Mid West is highlight by the distance between the circles, located at the geographic centroids.|
|Hexagon Tessellation| Each state is represented by a hexagon of equal size. The neighbouring states are easily contrasted, however the north east regions has been displaced from their geographic location. The sparsity of the population in the Mid West is highlight by the light yellow color, the age adjusted rate in Kentucky is the darkest and its neighbors are similar.|

@@

<!-- What is a contiguous cartogram -->
<!-- Intentionally preserve neighbors -->

@@

The groundwork for this approach had been laid in the mid 1930's by Raisz, and rectangular cartograms provide dramatic comparisons and are especially useful for correcting misconceptions communicated by geographic maps. 
@TFYCC quotes the official definition of Value-Area Cartograms, the simplistic displays which represent each area as a single rectangle, sized according to the value of the statistic.
This rectangular display also allows for tiling, where geographic neighbors placed in suitable relative positions also share borders, however contiguity may be sacrificed [@CDWCS].
Rectangular cartograms allow for bi-variate displays, population can be effectively communicated by the size of each rectangular, and a second variable can be communicated using color [@ORC].

@@

A simple tile map presents the areas in a tessellated grid display, where geographic neighbors are found next to each other, with some necessary displacement employed for regions with more than four neighbors.
These tiles may be labelled or colored to represent a value.
Tile maps may be difficult to create, they are best created manually, with additional time and care required as the amount of geographic areas to include increases.
@MDAC define the term 'mosaic cartograms' for hexagonal tile displays, where the number of tiles for each area can be used to communicate the statistic of regions. 
The complexity of the boundaries can be adjusted in the resulting display, as the size of the tiles used allows a trade-off be made between boundary complexity and simplicity.

@@

<!-- Choropleths, true topology, geographic distance, good for showing geographic importance -->
To choose an appropriate map display, the map creator must consider the intended user, and message the map will communicate.
Choropleth displays utilize more traditional cartographic methods, they are usually true to the topography, displaying familiar boundaries of countries, states or administrative areas. Highlighting the geographic distribution, and differences between equal spatial units.
When there is a relationship between the variable being mapped and geographical features, a choropleth map would pair this information visually.
Choropleth methods require a decision to be made about the projection of the display, in this case aim to find a map projection that gives the least distorted representation of the geography [@UMPT].
However a choropleth is limited to a uni-variate display, presenting a spatial distribution by filling each area with a color.
These displays favor geographically large regions over small regions, though inset displays are often used to zoom in on a geographic area containing many small regions, this is seen in Figure \ref{fig:choropleth-grid-create} c of the state of Queensland, where the capital city Brisbane is highlighted by the inset map. 
Users can identify the areas relevant to them by the familiar boundaries, and can easily contrast the experience of geographic neighbors. The map is reusable and fit for many purposes.

<!-- Cartograms, sacrifice geography, focus on the outliers/ relationships in data space -->
Cartographers were historically the creators of geographic displays, and epidemiologists utilized the maps that had been prepared in advance.
Cartograms with the population variable used to distortion the geographic map regions and boundaries are density equalising maps, that reduce the visual prominence of large, but low-population areas [@VSSDCUC]. 
Cartogram displays incorporate the statistics and population into the design of the display,
shifting and sometimes sacrificing familiar boundaries to draw attention to large outliers in the data space or population density.
The unusual shapes are not longer familiar and may be difficult to compare. This can prevent effective comparison of the statistic. Cartograms also favor extreme statistics in the positive direction. Small extreme values are often lost due to the resulting small sized areas the do not draw user's attention.
The change in the map based on one variable provides the opportunity for bi-variate displays, using the change in area for one variable, and color for another.
The difference between the familiar map display helps highlight the impact of the disease on communities.
